---
title: "PreCheck Donkey data - Judas efficiency"
author: "Carlo Pacioni"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Load libraries
```{r}
# Load libraries
options(java.parameters = "-Xmx8024m")
library(XLConnect, quietly = T)
library(ggplot2, quietly = T)
library(data.table, quietly = T)
```
Read data
```{r}
data.path <- "../Data/"
judas.master <- data.table(readWorksheetFromFile(
  file = file.path(data.path, "Tracking_History_Judas_with habitat_MZ.xlsx"), 
                      sheet="Judas_Tracking_History"))
names(judas.master)
```

Check formatting
```{r}
sapply(judas.master, class)
```

LONG is character while it should be numeric. Checked what values are not numeric.

```{r}
judas.master[is.na(sapply(judas.master[, LONG], as.numeric)), ]

```

```{r}
# Number dead
judas.master[, sum(ACTION == "DEAD")]

# Number collared
judas.master[, sum(ACTION == "COLLARED")]

# Number of judas in DB
judas.master[, length(unique(JUDAS_ID))]

# Rm useless spaces in REGION
judas.master[, REGION := sub(pattern = " +", replacement = "", x = REGION)]
# Number by region/shire
judas.master[, .(njudas=length(unique(JUDAS_ID)), 
                 start.date=min(EVENT_DATE), end.date=max(EVENT_DATE)), by=REGION]

judas.master[, .(njudas=length(unique(JUDAS_ID)), 
                 start.date=min(EVENT_DATE), end.date=max(EVENT_DATE)), 
                 by=c("REGION", "SHIRE")]
```
Check whether there are animals that moved between shires during the study (it seems none)
```{r}
# build a function where IDbyLoc is a data.table with one col reporting locations
   # (in this case shire), and the second, named 'IDs', is the judas' ids for each location
   # location is a character vector of with the heading of the column reporting the location
   # (here: location<- "SHIRE")
check.migration <- function(IDbyLoc, location) {
  areas <- unique(IDbyLoc[[location]])
  nareas <- length(areas)
  un.IDs <- IDbyLoc[, unique(IDs)]
  setkeyv(IDbyLoc, location)
  l2 <- vector(mode = "list", length = nareas)
  for (a in areas) {
    l2[[a]] <- vector(length = length(un.IDs))
    names(l2[[a]]) <- un.IDs
    for (u in un.IDs) {
      l2[[a]][u] <- sum(IDbyLoc[a, IDs] %in% un.IDs[u]) > 0
    }
  }
  names(l2) <- areas
  d <- do.call(cbind, args = l2)
  keep.row <- apply(d, MARGIN = 1, sum)
  d <- cbind(un.IDs, data.table(d))
  keep.row <- keep.row > 1
  return(d[keep.row,])
}

# Create IDbyLoc
IDbyShire <- judas.master[, .(IDs=unique(JUDAS_ID)), by=SHIRE]
check.migration(IDbyShire, location = "SHIRE")

```

Clean up and generating variables
```{r}
# Rm found dead because they are out of the program
judas.cleaned <- judas.master[ACTION != "DEAD", ]

# Rm not found because they do not contribute 
judas.cleaned <- judas.cleaned[ACTION != "NONE", ]

# Cross check totals are still the same
judas.cleaned[, .(njudas=length(unique(JUDAS_ID)), 
                 start.date=min(EVENT_DATE), end.date=max(EVENT_DATE)), 
              by=c("REGION", "SHIRE")]

# Rm judas with no long
keep <- !is.na(sapply(judas.cleaned[, LONG], as.numeric))
judas.cleaned <- judas.cleaned[keep, ]
judas.cleaned[is.na(sapply(judas.cleaned[, LONG], as.numeric)), ]
judas.cleaned[, LONG := as.numeric(LONG)]

# Set start and date
judas.cleaned[, start.date := min(EVENT_DATE), by=JUDAS_ID]
judas.cleaned[, end.date := max(EVENT_DATE), by=JUDAS_ID]

# Cross check start.date matches collared
judas.cleaned[ACTION == "COLLARED", date.coll := EVENT_DATE, by=JUDAS_ID]
judas.cleaned[ACTION == "COLLARED", sum(start.date != date.coll, na.rm = T)]
judas.cleaned[start.date != date.coll, ]

# Rm date.coll
judas.cleaned[, date.coll := NULL]

# Length in the program
judas.cleaned[, Time.deployment := difftime(end.date, start.date, units="weeks")]
judas.cleaned[, Time.dep.years := round(as.numeric(Time.deployment) / 52, 2)]

ggplot(judas.cleaned, aes(Time.dep.years)) + geom_histogram()
summary(judas.cleaned[, Time.dep.years])

```

Some animals have end.date set on the day they have been collared (listed below). I'm assumed that these were animals that were collared on start.date and searched and not found, is this correct?
Some other aniamls were tracked for several years, it is correct?

```{r}
# List of animals that have been deployed for 0 
judas.cleaned[Time.deployment == 0, .SD, 
              .SDcols=c("SHIRE", "JUDAS_ID", "EVENT_DATE", "EVENT_ID", "EVENT",
                        "ACTION")]

judas.cleaned <- judas.cleaned[Time.deployment>0,]

# List of animals that have been deployed for > 10 yrs 
judas.cleaned[Time.dep.years > 10,  unique(JUDAS_ID)]
```

Check whether there are judas with < 5 data points and rm
```{r}
locs <- judas.cleaned[, .N, by=JUDAS_ID]
locs[, sum(N<6)]
IDs.rm <- locs[N<6, JUDAS_ID]
judas.cleaned <- judas.cleaned[!JUDAS_ID %in% IDs.rm, ]
```


Save final dataset and csv with sample size
```{r}
dir.create(file.path(data.path, "Analysis"))
save(judas.cleaned, file = file.path(data.path, "judas.cleaned.rda"))
descr.fin <- judas.cleaned[, .(njudas=length(unique(JUDAS_ID)), 
                 start.date=min(EVENT_DATE), end.date=max(EVENT_DATE)), 
                 by=c("REGION", "SHIRE")]
descr.fin
descr.fin[, sum(njudas)]
write.csv(descr.fin, file = file.path(data.path, "Analysis", "descr.fin.csv"), 
          row.names = F)

ntrack.events <- judas.cleaned[, .N, by=JUDAS_ID]
ntrack.events[, summary(N)]
```