---
title: "PreCheck Donkey data - Judas efficiency"
author: "Carlo Pacioni"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Load libraries
```{r}
# Load libraries
options(java.parameters = "-Xmx8024m")
library(XLConnect, quietly = T)
library(ggplot2, quietly = T)
library(data.table, quietly = T)
```
Read data
```{r}
data.path <- "../Data/"
judas.master <- data.table(readWorksheetFromFile(
  file = file.path(data.path, "Tracking_History_Judas_with habitat_MZ.xlsx"), 
                      sheet="Judas_Tracking_History"))
names(judas.master)
```

Check formatting
```{r}
sapply(judas.master, class)
```

LONG is character while it should be numeric. Checked what values are not numeric.

```{r}
judas.master[is.na(sapply(judas.master[, LONG], as.numeric)), ]

```

```{r}
# Number dead
judas.master[, sum(ACTION == "DEAD")]

# Number collared
judas.master[, sum(ACTION == "COLLARED")]

# Number of judas in DB
judas.master[, length(unique(JUDAS_ID))]

# Rm useless spaces in REGION
judas.master[, REGION := sub(pattern = " +", replacement = "", x = REGION)]
# Number by region/shire
judas.master[, .(njudas=length(unique(JUDAS_ID)), 
                 start.date=min(EVENT_DATE), end.date=max(EVENT_DATE)), by=REGION]

judas.master[, .(njudas=length(unique(JUDAS_ID)), 
                 start.date=min(EVENT_DATE), end.date=max(EVENT_DATE)), 
                 by=c("REGION", "SHIRE")]
```
Check whether there are animals that moved between regions, shires or property during the study (it seems none, is this correct?)
```{r}
# build a function where IDbyLoc is a data.table with one col reporting locations
   # (e.g. shire), and the second, named 'IDs', is the judas' ids within each location
check.migration <- function(IDbyLoc, location) {
  areas <- unique(IDbyLoc[[location]])
  nareas <- length(areas)
  # Unique IDs
  un.IDs <- IDbyLoc[, unique(IDs)]
  names(un.IDs) <- IDbyLoc[, unique(IDs)]
  setkeyv(IDbyLoc, location)
  l2 <- vector(mode = "list", length = nareas)
  # For each area checks what ID are present
  for (a in areas) {
    l2[[a]] <- vector(length = length(un.IDs))
    names(l2[[a]]) <- un.IDs
    for (u in un.IDs) {
      l2[[a]][u] <- sum(IDbyLoc[a, IDs] %in% un.IDs[u]) > 0
    }
  }
  # Combined in a data.frame where locations are columns
  d <- do.call(cbind, args = l2)
  # sum row-wise, if the same ID is present in more than one location the sum>1
  keep.row <- apply(d, MARGIN = 1, sum)
  d <- cbind(un.IDs, data.table(d))
  keep.row <- keep.row > 1
  return(d[keep.row,])
}

# Create IDbyLoc
IDbyShire <- judas.master[, .(IDs=unique(JUDAS_ID)), by=SHIRE]
check.migration(IDbyShire, location = "SHIRE")

IDbyRegion <- judas.master[, .(IDs=unique(JUDAS_ID)), by=REGION]
check.migration(IDbyRegion, location = "REGION")

IDbyArea <- judas.master[, .(IDs=unique(JUDAS_ID)), by=AREA]
check.migration(IDbyArea, location = "AREA")

```

Clean up and generating variables
```{r}
# Rm found dead because they are out of the program
judas.cleaned <- judas.master[ACTION != "DEAD", ]

# Rm not found because they do not contribute 
judas.cleaned <- judas.cleaned[ACTION != "NONE", ]

# Cross check totals are still the same
judas.cleaned[, .(njudas=length(unique(JUDAS_ID)), 
                 start.date=min(EVENT_DATE), end.date=max(EVENT_DATE)), 
              by=c("REGION", "SHIRE")]

# Rm judas with no long
keep <- !is.na(sapply(judas.cleaned[, LONG], as.numeric))
judas.cleaned <- judas.cleaned[keep, ]
judas.cleaned[is.na(sapply(judas.cleaned[, LONG], as.numeric)), ]
judas.cleaned[, LONG := as.numeric(LONG)]

# Set start and end date
judas.cleaned[, start.date := min(EVENT_DATE), by=JUDAS_ID]
judas.cleaned[, end.date := max(EVENT_DATE), by=JUDAS_ID]

# Cross check start.date matches collared
judas.cleaned[ACTION == "COLLARED", date.coll := EVENT_DATE, by=JUDAS_ID]
judas.cleaned[ACTION == "COLLARED", sum(start.date != date.coll, na.rm = T)]

```

Some animals have been collared after their start date. Is this a collar replacement?
```{r}
judas.cleaned[start.date != date.coll, ]

# Rm date.coll
judas.cleaned[, date.coll := NULL]
```

Let's have a look at the length in the program for each donkey"
```{r}
# Length in the program
judas.cleaned[, Time.deployment := difftime(end.date, start.date, units="weeks")]
judas.cleaned[, Time.dep.years := round(as.numeric(Time.deployment) / 52, 2)]

ggplot(judas.cleaned, aes(Time.dep.years)) + geom_histogram()
summary(judas.cleaned[, Time.dep.years])

```

Some animals have end.date set on the day they have been collared (listed below). I'm assumed that these were animals that were collared on start.date and searched and not found, is this correct?
Some other aniamls were tracked for several years, it is correct?

```{r}
# List of animals that have been deployed for 0 
judas.cleaned[Time.deployment == 0, .SD, 
              .SDcols=c("SHIRE", "JUDAS_ID", "EVENT_DATE", "EVENT_ID", "EVENT",
                        "ACTION")]

judas.cleaned <- judas.cleaned[Time.deployment>0,]

# List of animals that have been deployed for > 10 yrs 
judas.cleaned[Time.dep.years > 10,  unique(JUDAS_ID)]
```

Check whether there are judas with < 5 data points and rm
```{r}
locs <- judas.cleaned[, .N, by=JUDAS_ID]
locs[, sum(N<6)]
IDs.rm <- locs[N<6, JUDAS_ID]
judas.cleaned <- judas.cleaned[!JUDAS_ID %in% IDs.rm, ]
```

Calculates the home range centre as arithmetic mean of coordinate and compute deviation from the centre on the X and Y
```{r}
# Home Range centres
judas.cleaned[, ':='(HRlat=mean(LAT), HRlong=mean(LONG)), by=JUDAS_ID]

calc.latlong.dist<- function(xy1,xy2)
{
  # uses spherical law of cosines to calculate distance between two lat/long
  # coordinates in decimal degrees
  R <- 6371 # Earths radius
  xy1 <- (pi * xy1)/180 # radians
  xy2 <- (pi * xy2)/180 
  D <- acos(sin(xy1[,1])*sin(xy2[,1]) + cos(xy1[,1])*cos(xy2[,1])*cos(xy2[,2]-xy1[,2]))  
  return(R*D)
}

# Calculate deviations from HRcentres 
judas.cleaned[, xdev:=calc.latlong.dist(judas.cleaned[, .(LAT, HRlong)],
                                        judas.cleaned[, .(HRlat, HRlong)])]
judas.cleaned[, ydev:=calc.latlong.dist(judas.cleaned[, .(HRlat, LONG)],
                                        judas.cleaned[, .(HRlat, HRlong)])]
judas.cleaned[, summary(xdev)]
judas.cleaned[, summary(ydev)]

```

Something is clearly wrong as some animals have moved ~ 2,000 km north or south and >12,000 km east or west.
Let's check what happened:
```{r}
judas.cleaned[ydev>12000,]
```

Okay some animals have coordinates 0 (somewhere in central Africa!). I guess that these were meant to be NA, removed for now and check again.
```{r}
judas.cleaned <- judas.cleaned[LAT < 0,]
judas.cleaned[, ':='(HRlat=mean(LAT), HRlong=mean(LONG)), by=JUDAS_ID]
judas.cleaned[, xdev:=calc.latlong.dist(judas.cleaned[, .(LAT, HRlong)],
                                        judas.cleaned[, .(HRlat, HRlong)])]
judas.cleaned[, ydev:=calc.latlong.dist(judas.cleaned[, .(HRlat, LONG)],
                                        judas.cleaned[, .(HRlat, HRlong)])]
judas.cleaned[, summary(xdev)]
judas.cleaned[, summary(ydev)]

```

This is better, but still the distibution is very odd... 75% moved less than 6 km but max value is 300 or 500 km?
```{r}
judas.cleaned[xdev > 200,]
judas.cleaned[, summary(LAT)]
```

Some donkey have LAT<-14 and are in the middle of the Timor sea! Other have done a sudden move of >200 km... is that possible?

For now I keep only entries where xdev and ydev is less than ~20 km
```{r}
judas.cleaned <- judas.cleaned[xdev < 20,]
judas.cleaned[, ':='(HRlat=mean(LAT), HRlong=mean(LONG)), by=JUDAS_ID]
judas.cleaned[, xdev:=calc.latlong.dist(judas.cleaned[, .(LAT, HRlong)],
                                        judas.cleaned[, .(HRlat, HRlong)])]
judas.cleaned[, ydev:=calc.latlong.dist(judas.cleaned[, .(HRlat, LONG)],
                                        judas.cleaned[, .(HRlat, HRlong)])]
judas.cleaned[, summary(xdev)]
judas.cleaned[, summary(ydev)]

judas.cleaned <- judas.cleaned[ydev < 20,]
judas.cleaned[, ':='(HRlat=mean(LAT), HRlong=mean(LONG)), by=JUDAS_ID]
judas.cleaned[, xdev:=calc.latlong.dist(judas.cleaned[, .(LAT, HRlong)],
                                        judas.cleaned[, .(HRlat, HRlong)])]
judas.cleaned[, ydev:=calc.latlong.dist(judas.cleaned[, .(HRlat, LONG)],
                                        judas.cleaned[, .(HRlat, HRlong)])]
judas.cleaned[, summary(xdev)]
judas.cleaned[, summary(ydev)]
```



Save final dataset and csv with sample size
```{r}
dir.create(file.path(data.path, "Analysis"))
save(judas.cleaned, file = file.path(data.path, "judas.cleaned.rda"))
descr.fin <- judas.cleaned[, .(njudas=length(unique(JUDAS_ID)), 
                 start.date=min(EVENT_DATE), end.date=max(EVENT_DATE)), 
                 by=c("REGION", "SHIRE")]
descr.fin
descr.fin[, sum(njudas)]
write.csv(descr.fin, file = file.path(data.path, "Analysis", "descr.fin.csv"), 
          row.names = F)

ntrack.events <- judas.cleaned[, .N, by=JUDAS_ID]
ntrack.events[, summary(N)]
```